Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-11-12 paper/appendix.tex
a
sonquist, new proof, end of 2.2

\begin{theorem*}[B2B consistency - general case]

     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centred and full rank noise.
     %
     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimize
     %
     $\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$and satisfy $E\hat H = \hat H$
\end{theorem*}

\begin{proof}

 Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B, we have
 \begin{align*}
    \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
%                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2]\\
                                                        &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
     \label{eq:doublenorm}
\end{align*}
\begin{align*}
    \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
%    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]\\
    &= EF \hat G
\end{align*}

Let us prove that $EF\hat G = F\hat G$, that is, that the $d_x$ bottom rows of $F\hat G$ are zero. Let $Z=F^\dagger EF\hat G$. We have $FZ = FF^\dagger EF  \hat G= EF\hat G$ ($FF^\dagger E =E$ since $F$ is full rank on $Img(E)$). Since E is a contraction, we have $ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$. Therefore,
 $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$

But as $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, the above inequality is an equality, $Z=\hat G$ and $EF\hat G = F\Hat G$. Therefore, $\left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 = \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 = \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2$ is minimised by $\hat G$ and $\hat H$. Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$. This completes the proof.
\end{proof}

@@

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_supp.pdf}
  \caption{Average $\Delta R$ for each $Y$ dimension (Forward model), canonical components (PLS and CCA) or feature (B2B) across subjects. Formally, Backward model cannot have a $\Delta R$ because it never demix multiple $X$ features. In such cases, we thus simply report the decoding score.}
  \label{fig:meg_supp}
\end{figure}

>>>>

\begin{theorem}[B2B consistency - general case]

     Consider the B2B model from Equation $$Y = (XE + N)F$$ $N$ centred and full rank noise.

     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimise

     $$\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$$ and satisfy

     $$E\hat H = \hat H$$
\end{theorem}
\begin{proof}

 Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B.

 Since $\hat G$ is the least square estimator of $X$ from $Y$
 \begin{align*}
    \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2]
\end{align*}
Replacing $Y$ by its model definition $Y = (XE+N)F$, we have
 \begin{align*}
    \hat G &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2] =\arg \min_G \mathbb{E}[\left \| X - XEFG + NFG \right\|^2]
  \end{align*}
  Since $N$ is centered and independent of $X$, we have
  \begin{align}
    	  \hat G &=  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
     \label{eq:Gdoublenorm}
\end{align}

Samely, for $\hat H$, we have
\begin{align*}
    \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]
 \end{align*}
 a positive quantity which reaches a minimum (zero) for
 \begin{align}
    \hat H = EF \hat G
    \label{eq:Hdoublenom}
\end{align}

Let us now prove that $EF\hat G = F\hat G$.

Let $F^\dagger$ be the pseudo inverse of $F$, and $Z=F^\dagger EF\hat G$, we have $FZ = FF^\dagger EF \hat G$

Since $F$ is full rank on $Img(E)$, we have $FF^\dagger E =E$, and $FZ = EF\hat G$

As $E$ is a binary diagonal matrix, it is an orthogonal projection and therefore a contraction, thus
 $$ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$$ and
 $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$

But since $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, we also have
$$\left \| X - XEF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 \leq \left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2$$

Summarizing the above,
$$\left \| X - XEF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
$$\left \| X - XEF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2$$
$$\left \| NF\hat G\right \| ^2 =  \| NEF\hat G \| ^2$$

$N$ being full rank, this yields $EF\hat G = F\Hat G$.

Replacing into $\eqref{eq:Gdoublenorm}$, and setting $H = EFG$, we have
\begin{align*}
	\hat G &=  \arg \min_G  \left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 \\
	&=   \arg \min_G \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 \\
	\hat H &=  \arg \min_H \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2
	\label{eq:4}
\end{align*}

Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$, since $E$, a binary diagonal matrix, is involutive. This completes the proof.
\end{proof}

@@

\subsection{Binarizing $\hat{H}$ - the Sonquist-Morgan criterion}

Back to back regression provides us with $Diag(\hat{H})$, a vector of real numbers that estimate the causal influence of features in $X$. In order to derive the true causal influence matrix $E$ , a binary matrix, we need to binarize  $\hat{H}$, i.e. test each feature for equality with zero. In regression analysis, the traditional approach to this problem employs a $t$-test to check whether the regression coefficients differ from zero \citep{student1908probable}. This test will not succeed here: $\hat{H}$ is obtained from a double regression with regularization, and will not follow the t-distribution.

Instead, we treat the binarization of $\hat{H}$ as a clustering problem: separate the elements in the diagonal into a group of ``small values'', and a group of ``large values''. More specifically, we propose to maximize the ratio of inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
%
Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan \citep{sonquist_morgan} \citep{kass75} criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
%
To binarize $\hat{H}$, set to one all the $p$ diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the $d_x-p$ diagonal entries.

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_supp.pdf}
  \caption{Average $\Delta R$ for each $Y$ dimension (Forward model), canonical components (PLS and CCA) or feature (B2B) across subjects. Formally, Backward model cannot have a $\Delta R$ because it never unmixes the multiple $X$ features. In such cases, we thus simply report the decoding score.}
  \label{fig:meg_supp}
\end{figure}

### 2019-11-12 paper/paper.tex
sonquist, new proof, end of 2.2

     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centred and full rank noise.
     %
     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimize
     %
     $\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$and satisfy $E\hat H = \hat H$
\end{theorem}
%
\begin{proof}
  See Appendix \ref{appendix:theorem_proof}.
\end{proof}
%
% \begin{proof}
%
%  Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B, we have
%  \begin{align*}
%     \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
% %                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2]\\
%                                                         &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
%      \label{eq:doublenorm}
% \end{align*}
% \begin{align*}
%     \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
%     &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
% %    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]\\
%     &= EF \hat G
% \end{align*}
%
% Let us prove that $EF\hat G = F\hat G$, that is, that the $d_x$ bottom rows of $F\hat G$ are zero. Let $Z=F^\dagger EF\hat G$. We have $FZ = FF^\dagger EF  \hat G= EF\hat G$ ($FF^\dagger E =E$ since $F$ is full rank on $Img(E)$). Since E is a contraction, we have $ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$. Therefore,
%  $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
%
% But as $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, the above inequality is an equality, $Z=\hat G$ and $EF\hat G = F\Hat G$. Therefore, $\left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 = \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 = \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2$ is minimised by $\hat G$ and $\hat H$. Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$. This completes the proof.
% \end{proof}
%
\begin{equation}
  \text{So, }\hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top.
\end{equation}
% So, $\hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top$.

Assuming, without loss of generality, that the active features are the $k$ first, we have

% $$
\begin{equation}
  X^\top X = \left(\begin{array}{lccl}\Sigma_{X1} & \Sigma_{X2} \\ \Sigma_{X2} & \Sigma_{X3}\end{array}\right),\qquad\qquad N^\top N = \left(\begin{array}{lccl}\Sigma_{N1} & \Sigma_{N2} \\ \Sigma_{N2} & \Sigma_{N3}\end{array}\right),
\end{equation}
% $$
% and
\begin{equation}
  \hat H = \left(\begin{array}{cc}(\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1} & (\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X2} \\0 & 0\end{array}\right),
\end{equation}
\begin{equation}
  Diag_k (\hat H) = Diag((\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1}) = Diag((I+\Sigma_{X1}^{-1}\Sigma_{N1})^{-1}).
\end{equation}
%
% $$
% $$
% $$
% $$
%
In the absence of noise, we have $Diag(\hat H) = Diag(E)$. Otherwise the $k$ first diagonal elements of $\hat H$ are all positive, and bounded by $\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}}$ and $\frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$, where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X1}$ and $\Sigma_{N1}$.

The average value of the coefficients of $Diag(\hat H)$ is the trace of $\hat H$ divided by $k$. On average, since $X$ and $N$ are not correlated by definition, $\Sigma_{X1}^{-1}\Sigma_{N1}$ should have a low condition number, and $\frac{Var(X)}{Var(X)+Var(N)}$ provides a good estimate of $\frac{1}{k}Tr(\hat H)$.

>>>>

     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centred and full rank noise.
     %
     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimize
     %
     $\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$and satisfy $E\hat H = \hat H$
\end{theorem}
%
\begin{proof}
  See Appendix \ref{appendix:theorem_proof}.
\end{proof}

Since  $E\hat H = \hat H$, we have
\begin{equation}
  \hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top.
\end{equation}
% So, $\hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top$.

Assuming, without loss of generality, that the active features in $E$ are the $k$ first, and rewriting $X=(X_1,X_2)$ and $N=(N_1,N_2)$ ($X_1$ and $N_1$ containing the $k$ first features), we have ($\Sigma_{A B}$ denoting the covariance of $A$ and $B$)

% $$
\begin{equation}
  X^\top X = \left(\begin{array}{lccl}\Sigma_{X_1 X_1} & \Sigma_{X_1 X_2} \\ \Sigma_{X_1 X_2} & \Sigma_{X_2 X_2}\end{array}\right),\qquad\qquad N^\top N = \left(\begin{array}{lccl}\Sigma_{N_1 N_1} & \Sigma_{N_1 N_2} \\ \Sigma_{N_1 N_2} & \Sigma_{N_2 N_2}\end{array}\right),
\end{equation}
% $$
% and
\begin{equation}
  \hat H = \left(\begin{array}{cc}(\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1} & (\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_2} \\0 & 0\end{array}\right),
\end{equation}
\begin{equation}
  Diag_k (\hat H) = Diag((\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1}) = Diag((I+\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1})^{-1}).
\end{equation}
%
% $$
% $$
% $$
% $$
%
In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so $Diag_k(\hat H)=I$, and $$Diag(\hat H) = Diag(E)$$

Otherwise the $k$ first diagonal elements of $\hat H$ are all positive (since we assume that $F$ has full rank over $Img(E)$, they cannot be zero), lower than $1$, and bounded by 
$$\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}} \text{ and }  \frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$$where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X_1 X_1}$ and $\Sigma_{N_1 N_1}$.
In real cases, noise will be added to the diagonal elements of $\hat H$, by non-linearities and sampling error.

The average value of the coefficients of $Diag_k(\hat H)$ is the trace of $\hat H$ divided by $k$. Since $X$ and $N$ are not correlated, $\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1}$ should have a low condition number, and $\frac{Var(X)}{Var(X)+Var(N)}$ provides a good estimate of $\frac{1}{k}Tr(\hat H)$, that could be used to test whether a feature of $X$ has causal influence. The Sonquist and Morgan criterion defined in the appendix is another possibility.
